---
title: "SOM - Read and data plotting from the manuscript Paixao et al. 2021 - JASr"
author: "Paixão et al.2021 - JARs"
date: "28/09/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../", comment = NA, message = FALSE, indent = "", error = TRUE)

```

__Brief description of the script__

This R markdown document reads, summarizes and plots data for the manuscript *Paixão et al. 2021 Using Mechanical experiments to study Ground Stone Tool use: exploring the formation of percussive and grinding wear traces on Limestone tools. Journal of Archaeological Science: Reports*


The document contains:

1. Manuscript tables
2. Manuscript figures (data analysis)
3. Supplementary material, including extra tables and figures (data analysis)

This R project and respective scripts follow the procedures described by Marwick et al. 2017. 

To compile this markdown document do not delete or move files from their original folders.
Please note that most of the tables and figures in this file do not match the numbering in the original manuscript.

For any questions, comments and inputs, please contact:

João Marreiros, marreiros@rgzm.de, or Eduardo Paixão, paixao@rgzm.de

# Load data into R project

*Imported files are in: '`r paste0("../analysis/raw_data")`'*

*Figures are saved in: '`r paste0("../analysis/plots")`'*

*Tables are saved in: '`r paste0("../analysis/derived_data")`'*

## Load libraries

```{r}

library(tidyverse)
library(utils)
library(knitr)
library(janitor)
library(kableExtra)
library(GGally)
library(doBy)
library(ggpubr)
library(ggfortify)
library(tools)

```

## Import datasets

```{r}

gisdata <- read_csv("../raw_data/gisdata.csv")
confocaldata <- read.csv("../raw_data/confocaldata.csv", na.strings = "*****", encoding = "UTF-8")

data_file <- list.files("../raw_data/", pattern = "\\.csv$", full.names = TRUE)
md5_in <- md5sum(data_file)
info_in <- data.frame(file = basename(names(md5_in)), checksum = md5_in, row.names = NULL)

```


In this study two datasets are used:

1) __gisdata.csv__: dataset for the QGIS analysis
```{r}

str(gisdata)

```

2) __confocaldata.csv__: dataset for the Confocal microscopy surface texture analysis analysis
```{r}

str(confocaldata)

```

\newpage

# GIS analysis, Terrain analysis for Slope and TRI based on the 3D surface point clouds
## Slope

```{r fig.height = 6.5, fig.width = 8.5, res = 300}

# Compute proportions for perimeter and area grouped by sample and GIS parameter

slope <- filter(gisdata, parameter == "slope")
slopebefore <- filter(slope, cycle == "before")
slopeafter <-  filter(slope, cycle == "after")

# before experimental cycles (i.e. natural surfaces)

id2.5before <- filter(slopebefore, sample == "id2-5")
id3.3before <- filter(slopebefore, sample == "id3-3")
id3.8before <- filter(slopebefore, sample == "id3-8")
id3.9before <- filter(slopebefore, sample == "id3-9")
id6.1before <- filter(slopebefore, sample == "id6-1")
id6.3before <- filter(slopebefore, sample == "id6-3")
id6.6before <- filter(slopebefore, sample == "id6-6")
id6.7before <- filter(slopebefore, sample == "id6-7")

id2.5before <- id2.5before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.3before <- id3.3before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.8before <- id3.8before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.9before <- id3.9before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.1before <- id6.1before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.3before <- id6.3before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.6before <- id6.6before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.7before <- id6.7before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

# after experimental cycles

id2.5after <- filter(slopeafter, sample == "id2-5")
id3.3after <- filter(slopeafter, sample == "id3-3")
id3.8after <- filter(slopeafter, sample == "id3-8")
id3.9after <- filter(slopeafter, sample == "id3-9")
id6.1after <- filter(slopeafter, sample == "id6-1")
id6.3after <- filter(slopeafter, sample == "id6-3")
id6.6after <- filter(slopeafter, sample == "id6-6")
id6.7after <- filter(slopeafter, sample == "id6-7")

id2.5after <- id2.5after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.3after <- id3.3after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.8after <- id3.8after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.9after <- id3.9after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.1after <- id6.1after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.3after <- id6.3after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.6after <- id6.6after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.7after <- id6.7after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)


newslope <- do.call("rbind", list(id2.5before, id3.3before, id3.8before, id3.9before, id6.1before, id6.3before, id6.6before, id6.7before, id2.5after, id3.3after, id3.8after, id3.9after, id6.1after, id6.3after, id6.6after, id6.7after))

# save outputs

write_csv(newslope,"../derived_data/newslope.csv")

# Plot data

# Area %

impactdf <- filter(newslope, motion == "Impact")
grinding <- filter(newslope, motion == "Grinding")


areaimpact <- ggplot(impactdf, aes(x = elev_max, y = areaperc, colour = cycle)) +
  geom_line() +
  facet_wrap(~sample, scale = "free") +
  ggtitle("Slope analysis, impact") +
  ylab("Area %") +
  xlab("Slope in º")

areaimpact

ggsave("../plots/slopeareaimpact.png")


areagrinding <- ggplot(grinding, aes(x = elev_max, y = areaperc, colour = cycle)) +
  geom_line() +
  facet_wrap(~sample, scale = "free") +
  ggtitle("Slope analysis, grinding") +
  ylab("Area %") +
  xlab("Slope in º")

areagrinding

ggsave("../plots/slopeareagrinding.png")

# Perimeter %

perimimpact <- ggplot(impactdf, aes(x = elev_max, y = perimperc, colour = cycle)) +
  geom_line() +
  facet_wrap(~sample) +
  ggtitle("Slope, impact") +
  ylab("Perimeter %") +
  xlab("Slope in º")

perimimpact

ggsave("../plots/slopeperimimpact.png")

perimgrinding <- ggplot(grinding, aes(x = elev_max, y = perimperc, colour = cycle)) +
  geom_line() +
  facet_wrap(~sample, scale = "free") +
  ggtitle("Slope, grinding") +
  ylab("Peerimeter %") +
  xlab("Slope in º")

perimgrinding

ggsave("../plots/slopeperimgrinding.png")


```

## TRI (Terrain roughness index)

```{r fig.height = 6.5, fig.width = 8.5, res = 300}

tri <- filter(gisdata, parameter == "tri")
tribefore <- filter(tri, cycle == "before")
triafter <- filter(tri, cycle =="after")


# before experimental cycles (i.e. natural surfaces)

id2.5before <- filter(tribefore, sample == "id2-5")
id3.3before <- filter(tribefore, sample == "id3-3")
id3.8before <- filter(tribefore, sample == "id3-8")
id3.9before <- filter(tribefore, sample == "id3-9")
id6.1before <- filter(tribefore, sample == "id6-1")
id6.3before <- filter(tribefore, sample == "id6-3")
id6.6before <- filter(tribefore, sample == "id6-6")
id6.7before <- filter(tribefore, sample == "id6-7")

id2.5before <- id2.5before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.3before <- id3.3before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.8before <- id3.8before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.9before <- id3.9before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.1before <- id6.1before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.3before <- id6.3before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.6before <- id6.6before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.7before <- id6.7before %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)


# after experimental cycles

id2.5after <- filter(triafter, sample == "id2-5")
id3.3after <- filter(triafter, sample == "id3-3")
id3.8after <- filter(triafter, sample == "id3-8")
id3.9after <- filter(triafter, sample == "id3-9")
id6.1after <- filter(triafter, sample == "id6-1")
id6.3after <- filter(triafter, sample == "id6-3")
id6.6after <- filter(triafter, sample == "id6-6")
id6.7after <- filter(triafter, sample == "id6-7")

id2.5after <- id2.5after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.3after <- id3.3after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.8after <- id3.8after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id3.9after <- id3.9after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.1after <- id6.1after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.3after <- id6.3after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.6after <- id6.6after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)

id6.7after <- id6.7after %>%
  group_by(sample) %>%
  mutate(
    areaperc = area / sum(area) * 100,
    perimperc = perimeter / sum(perimeter) * 100)


newtri <- do.call("rbind", list(id2.5before, id3.3before, id3.8before, id3.9before, id6.1before, id6.3before, id6.6before, id6.7before, id2.5after, id3.3after, id3.8after, id3.9after, id6.1after, id6.3after, id6.6after, id6.7after))

# save outputs

write_csv(newtri,"../derived_data/newtri.csv")

# Plot data

# Area %

impactdf <- filter(newtri, motion == "Impact")
grinding <- filter(newtri, motion == "Grinding")

areaimpact <- ggplot(impactdf, aes(x = elev_max, y = areaperc, colour = cycle)) +
  geom_line() +
  facet_wrap(~sample, scale = "free") +
  ggtitle("TRI analysis, impact") +
  ylab("Area %") +
  xlab("TRI")

areaimpact 

ggsave("../plots/triareaimpact.png")

areagrinding <- ggplot(grinding, aes(x = elev_max, y = areaperc, colour = cycle)) +
  geom_line() +
  facet_wrap(~sample, scale = "free") +
  ggtitle("TRI analysis, grinding") +
  ylab("Area %") +
  xlab("TRI")

areagrinding

ggsave("../plots/triareagrinding.png")

# Perimeter %

perimimpact <- ggplot(impactdf, aes(x = elev_max, y = perimperc, colour = cycle)) +
  geom_line() +
  facet_wrap(~sample) +
  ggtitle("TRI analysis, impact") +
  ylab("Perimeter %") +
  xlab("TRI")

perimimpact

ggsave("../plots/triperimimpact.png")

perimgrinding <- ggplot(grinding, aes(x = elev_max, y = perimperc, colour = cycle)) +
  geom_line() +
  facet_wrap(~sample, scale = "free") +
  ggtitle("TRI analysis, grinding") +
  ylab("Perimeter %") +
  xlab("TRI")

perimgrinding

ggsave("../plots/triperimgrinding.png")


```


# Confocal micro surface texture data

## Import and summarize data

```{r fig.height = 6.5, fig.width = 8.5, res = 300}

# compute descriptive statistics

nminmaxmeanmedsd <- function(x){
	y <- x[!is.na(x)]
	n_test <- length(y)
	min_test <- min(y)
	max_test <- max(y)
	mean_test <- mean(y)
 	med_test <- median(y)
 	sd_test <- sd(y)
 	out <- c(n_test, min_test, max_test, mean_test, med_test, sd_test)
 	names(out) <- c("n", "min", "max", "mean", "median", "sd")
 	return(out)
}

num.var <- 21:length(confocaldata)

confostats <- summaryBy(.~sample + motion + workedmaterial, data=confocaldata[c("sample", "motion","workedmaterial", names(confocaldata)[num.var])], FUN=nminmaxmeanmedsd)

write_csv(confostats, "../derived_data/confostats.csv")

```

## Plot all paramaters

```{r}

# Loop for plotting all surface texture parameters

for (i in num.var) cat("[",i,"] ", names(confocaldata)[i], "\n", sep = "")
	
for (i in num.var) {
  p <- ggplot(data = confocaldata, aes_string(x = "workedmaterial", y = names(confocaldata)[i],
	                                        colour = "motion")) + 
	     geom_boxplot() +
	     # geom_line(aes(group = motion)) +
	     theme_classic() +
	     labs(colour = "motion") +
	     # facet_wrap(~ sample) +
	     labs(x = "material", y = gsub("\\.", " ", names(confocaldata)[i])) +
	     scale_colour_hue(h = c(25,225), limits = levels(confocaldata[["motion"]]))
  print(p)
  
  # saves the plots 
  file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_plot_",
	                   names(confocaldata)[i], ".pdf")
	ggsave(filename = file_out, plot = p, path = "../plots", device = "pdf", width = 26,
	       height = 21, units = "cm" )
}

```

## Scatterplots of selected variables combined by Worked material and Motion

```{r}

# Sa vs. Sq

Sa_Sq <- ggplot(data = confocaldata) +  
         geom_point(mapping = aes(x = Sa, y = Sq, colour = workedmaterial)) +
         theme_classic() +
         labs(colour = "workedmaterial") +
         facet_wrap(~ motion) +
         scale_colour_hue(h = c(25, 230)) 
print(Sa_Sq)
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_scatterplot_Sa-Sq", ".pdf")
ggsave(filename = file_out, plot = Sa_Sq, path = "../plots", device = "pdf")

# epLsar vs. Asfc

ep_As <- ggplot(data = confocaldata) +  
         geom_point(mapping = aes(x = Fractal.complexity.Asfc, y = Lengthscale.anisotropy.Sfrax.epLsar, colour = workedmaterial)) +
         theme_classic() +
         labs(colour = "workedmaterial") +
         facet_wrap(~ motion) +
         scale_colour_hue(h = c(25, 230)) 
print(ep_As)
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_scatterplot_Asfc-epLsar", ".pdf")
ggsave(filename = file_out, plot = ep_As, path = "../plots", device = "pdf")

# Sq vs. Vmc

Sq_Vmc <- ggplot(data = confocaldata) +  
          geom_point(mapping = aes(x = Sq, y = Vmc, colour = workedmaterial)) +
          theme_classic() +
          labs(colour = "workedmaterial") +
          facet_wrap(~ motion) +
          scale_colour_hue(h = c(25, 230))  
print(Sq_Vmc)
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_scatterplot_Sq-Vmc", ".pdf")
ggsave(filename = file_out, plot = Sq_Vmc, path = "../plots", device = "pdf")

# Mean depth of furrows vs. mean density of furrows

furrows <- ggplot(data = confocaldata) +  
           geom_point(mapping = aes(x = Mean.depth.of.furrows, y = Mean.density.of.furrows,
                                    colour = workedmaterial)) +
           theme_classic() +
           labs(colour = "workedmaterial", x = "Mean depth of furrows", y = "Mean density of furrows") +
           facet_wrap(~ motion) +
           scale_colour_hue(h = c(25, 230))
print(furrows)
file_out <- paste0(file_path_sans_ext(info_in[["file"]]), "_scatterplot_furrows", ".pdf")
ggsave(filename = file_out, plot = furrows, path = "../plots", device = "pdf")

# combine all in a single image

ggarrange(Sa_Sq, Sq_Vmc, furrows, ep_As, common.legend = TRUE, legend = "bottom")

ggsave("../plots/scatterplots.png")

```

## Scatterplot matrix for the ISO 25178 Area scale, Height and volume parameters

```{r}

data(confocaldata, package = "reshape")

# Height parameters

ggpairs(data=confocaldata,
        columns = c(21:26),
        cardinality_threshold = 30,
        mapping = ggplot2::aes(color = workedmaterial, shape = motion),
        lower = list(continuous = wrap("points", alpha = 1, size = 2)),
        upper = list(continuous = "blank"),
        legend = c(2,1)
        ) +

    theme(legend.position = "right") +
  labs(fill = "Micro polish type")

ggsave("../plots/confocalarea_matrix.png")

# Volume parameters

ggpairs(data=confocaldata,
        columns = c(36:41),
        cardinality_threshold = 30,
        mapping = ggplot2::aes(color = workedmaterial, shape = motion),
        lower = list(continuous = wrap("points", alpha = 1, size = 2)),
        upper = list(continuous = "blank"),
        legend = c(2,1)
        ) +

    theme(legend.position = "right") +
  labs(fill = "Micro polish type")

ggsave("../plots/confocalvolume_matrix.png")

```

## Plot confostats for the ISO 25178 Area-scale, Height and volume parameters

```{r}
# select parameter from dataset

# first Height parameters

heightconfostats <- select(confostats,sample,workedmaterial, Sq.mean,Ssk.mean,Sku.mean,Sp.mean,Sv.mean,Sz.mean,Sa.mean)

p1 <- ggplot(heightconfostats, aes(x=workedmaterial, y=Sq.mean, colour=workedmaterial)) + 
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p2 <- ggplot(heightconfostats, aes(x=workedmaterial, y=Ssk.mean, colour=workedmaterial)) +   
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p3 <- ggplot(heightconfostats, aes(x=workedmaterial, y=Sku.mean, colour=workedmaterial)) +   
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p4 <- ggplot(heightconfostats, aes(x=workedmaterial, y=Sp.mean, colour=workedmaterial)) + 
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p5 <- ggplot(heightconfostats, aes(x=workedmaterial, y=Sv.mean, colour=workedmaterial)) + 
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p6 <- ggplot(heightconfostats, aes(x=workedmaterial, y=Sz.mean, colour=workedmaterial)) + 
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p7 <- ggplot(heightconfostats, aes(x=workedmaterial, y=Sa.mean, colour=workedmaterial)) + 
  geom_boxplot() +
  labs(x="", colour="Micro polish")

ggarrange(p1, p2, p3, p4, p5, p6, p7, common.legend = TRUE, font.label = list(size=8), legend="bottom")

ggsave("../plots/confostatsarea_boxplots.png")

# Now Volume parameters

volumeconfostats <- select(confostats,sample,workedmaterial, Vm.mean,Vv.mean,Vmp.mean,Vmc.mean,Vvc.mean,Vvv.mean)

p8 <- ggplot(volumeconfostats, aes(x=workedmaterial, y=Vm.mean, colour=workedmaterial)) + 
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p9 <- ggplot(volumeconfostats, aes(x=workedmaterial, y=Vv.mean, colour=workedmaterial)) +   
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p10 <- ggplot(volumeconfostats, aes(x=workedmaterial, y=Vmp.mean, colour=workedmaterial)) +   
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p11 <- ggplot(volumeconfostats, aes(x=workedmaterial, y=Vmc.mean, colour=workedmaterial)) + 
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p12 <- ggplot(volumeconfostats, aes(x=workedmaterial, y=Vvc.mean, colour=workedmaterial)) + 
  geom_boxplot() +
  labs(x="", colour="Micro polish")

p13 <- ggplot(volumeconfostats, aes(x=workedmaterial, y=Vvv.mean, colour=workedmaterial)) + 
  geom_boxplot() +
  labs(x="", colour="Micro polish")

ggarrange(p8, p9, p10, p11, p12, p13, common.legend = TRUE, font.label = list(size=8), legend="bottom")

ggsave("../plots/confostatvolume_boxplots.png")

```

# Show plot files information
```{r}
info_out <- list.files(path = "derived_data", pattern = "\\.pdf$", full.names = TRUE) %>% 
            md5sum()
```

# The checksum (MD5 hashes) of the exported files are:  
```{r, echo = FALSE}
info_out
```

# End of script
```{r}
sessionInfo()
```


